---
title: "10X Genomics analysis of ITP samples"
author: "David Schmidt, d.schmidt@sanquin.nl"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F,message = F)
```

```{r libs}
library(tidyverse)
library(Seurat)
library(future)
library(cowplot)
library(knitr)
library(beepr)
library(magrittr)
library(ggsci)
theme_set(theme_cowplot())

```

```{r load-data, eval=FALSE, include=FALSE}

## import 10x filtered barcode matrices and annotate
# add here later: annotation of TCR/BCR sequences


g001 = Read10X("data/count_matrices/SAN-DS-G001-filtered-feature-bc-matrix/") %>%
  CreateSeuratObject(counts=., project="T244")

g001 = AddMetaData(g001,
                      metadata=rep("chronic",length(colnames(g001))),
                      col.name="group")

g001 = AddMetaData(g001,
                      metadata=rep("T244",length(colnames(g001))),
                      col.name="pid")


g002 = Read10X("data/count_matrices/SAN-DS-G002-filtered-feature-bc-matrix/") %>%
  CreateSeuratObject(counts=., project="T263")

g002 = AddMetaData(g002,
                      metadata=rep("transient",length(colnames(g002))),
                      col.name="group")

g002 = AddMetaData(g002,
                      metadata=rep("T263",length(colnames(g002))),
                      col.name="pid")


g004 = Read10X("data/count_matrices/SAN-DS-G004-filtered-feature-bc-matrix/") %>%
  CreateSeuratObject(counts=., project="T281")

g004 = AddMetaData(g004,
                      metadata=rep("transient",length(colnames(g004))),
                      col.name="group")

g004 = AddMetaData(g004,
                      metadata=rep("T281",length(colnames(g004))),
                      col.name="pid")


g005= Read10X("data/count_matrices/SAN-DS-G005-filtered-feature-bc-matrix/") %>%
  CreateSeuratObject(counts=.,  project="T316")

g005 = AddMetaData(g005,
                      metadata=rep("T316",length(colnames(g005))),
                      col.name="pid")

g005 = AddMetaData(g005,
                      metadata=rep("chronic",length(colnames(g005))),
                      col.name="group")


g006 = Read10X("data/count_matrices/SAN-DS-G006-filtered-feature-bc-matrix/") %>%
  CreateSeuratObject(counts=.,project="T269")

g006 = AddMetaData(g006,
                      metadata=rep("T269",length(colnames(g006))),
                      col.name="pid")

g006 = AddMetaData(g006,
                      metadata=rep("transient",length(colnames(g006))),
                      col.name="group")


g008 = Read10X("data/count_matrices/SAN-DS-G008-filtered-feature-bc-matrix/") %>%
  CreateSeuratObject(counts=.,project="T440")

g008 = AddMetaData(g008,
                      metadata=rep("T440",length(colnames(g008))),
                      col.name="pid")

g008 = AddMetaData(g008,
                      metadata=rep("transient",length(colnames(g008))),
                      col.name="group")


# merge
ITPcells = merge(g001,y=c(g002,g004,g005,g006,g008), 
             add.cell.ids =  c("T244","T263","T281","T316","T269", "T440"),
             project="ITP")

saveRDS(ITPcells,"data/processed/20200625 allCells_beforeFiltering.rds")
rm(g001,g002,g004,g005,g006,g008,ITPcells)
```

**Data**

* scRNA-seq 
* scTCR-seq
* scBCR-seq

**Patients** 

| ID |	pid |	allocation |	age |	DiseaseCourse |	FirstCR |	IVIGResponse |
| --- | --- | --- | --- | --- | --- | --- |
| G001 |	T244 |	IVIG |	8,0 |	chronic |	never |	NR |
| G002 |	T263 |	Observation |	6,4 |	transient	| week1	| |
| G004 |	T281 | IVIG |	4,4 |	CSR |	week1 |	CSR |
| G005 |	T316 | Observation |	6,7 |	chronic |	never	| |
| G006	| T269 | Observation |	6,0 |	transient |	month1	| |
| G008	| T440 | IVIG |	4,5 |	CSR |	week1 |	CSR |

Where CSR = transient

```{r init}
allCells = readRDS("data/processed/20200625 allCells_beforeFiltering.rds")
table(allCells$pid) %>% kable(caption="Cells per subject before filtering")
table(allCells$group) %>% kable(caption="Cells per group before filtering")
```

```{r filtering-mt}
# calc mitochondrial QC metrics
allCells[["percent.mt"]] <- PercentageFeatureSet(object = allCells, pattern = "^MT-")
```

```{r qc, fig.height=13, fig.width=7}
# vis cell UMI / mt
VlnPlot(object = allCells,features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 1)
```

```{r qc-2, fig.height=7, fig.width=13}
plot1 = FeatureScatter(object = allCells, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot1 = plot1 +   geom_hline(yintercept=10)
plot2 = FeatureScatter(object = allCells, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot2 = plot2 + geom_hline(yintercept=c(300,2500))
CombinePlots(plots = list(plot1,plot2))
rm(plot1,plot2)
```

**Filtering**

* percent mtDNA < 0.1
* `nFeature_RNA` > 300 and < 2500

```{r include=F}
# filter data and save
allCells.filtered = subset(x = allCells, subset = nFeature_RNA > 300 & nFeature_RNA < 2500 & percent.mt < 10)

saveRDS(allCells.filtered,"data/processed/20200625 allCells_filtered.rds")
rm(allCells,allCells.filtered)
```

```{r init-2}
allCells = readRDS("data/processed/20200625 allCells_filtered.rds")
table(allCells$pid) %>% kable(caption="Cells per subject after filtering")
table(allCells$group) %>% kable(caption="Cells per group after filtering")
```

**Pre-processing** 

* Log normalization using `NormalizeData`. Feature counts for each cell are divided by the total counts for that cell and multiplied by the scale.factor. This is then natural-log transformed using log1p.
* finding top 2000 variable features using `FindVariableFeatures` with selection by vst 
* integration by CCA using first 20 dimensions
* scaling

```{r eval=FALSE, include=F}
# normalize and find variable features
allCells.list = SplitObject(allCells)
allCells.list <- lapply(X = allCells.list, FUN = function(x) {
    x <- NormalizeData(x)
    x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 2000)
})

# integrate by CCA with 20 dimensions
allCells.anchors <- FindIntegrationAnchors(object.list = allCells.list, dims = 1:20)
allCells <- IntegrateData(anchorset = allCells.anchors, dims = 1:20)

# scale data 
DefaultAssay(allCells) <- "integrated" 
allCells <- ScaleData(allCells) 
rm(allCells.anchors,allCells.list)

saveRDS(allCells,"data/processed/20200625 allCells_NormalizedIntegratedScaled.rds")
rm(allCells)
```

**Clustering** 

Elbow plot to determine optimal number of PCs

```{r}
# Determine elbow of PC plot 
allCells = readRDS("data/processed/20200625 allCells_NormalizedIntegratedScaled.rds")
allCells <- RunPCA(allCells, npcs = 50)
ElbowPlot(allCells,ndims = 50)
```

* Compute and store 13 PCs (elbow) by PCA using `RunPCA` using variable features for the Assay
* Dimensional reduction by UMAP using `RunUMAP` using PCA dimensional reduction as input, using the python package umap-learn
* Shared Nearest Neighbor (SNN) graph, determining k-nearest neighbors of each cell (k=20) using the 13 PC dimensions as input with `FindNeighbors`
* Identify clusters based on PCs, using the SNN, with the function `FindClusters` and a resolution of 0.8 (standard). The output is saved as `$seurat_clusters`

```{r cluster, eval=FALSE, include=F}
allCells <- RunPCA(allCells, npcs = 13) #recalc so we store only 13
allCells <- RunUMAP(allCells, reduction = "pca", dims = 1:13)
allCells <- FindNeighbors(allCells, reduction = "pca", dims = 1:13)
allCells <- FindClusters(allCells)

saveRDS(allCells,"data/processed/20200626 allCells_Clustered.rds")
rm(allCells)
```

Cells per cluster

### data load anchor

```{r}
allCells = readRDS("data/processed/20200626 allCells_Clustered.rds")
table(allCells$seurat_clusters) %>% kable(caption="Cells per identified seurat cluster(all samples)")

allClusters = table(allCells$seurat_clusters)  %>% as.data.frame() %>%
  rename( Cluster = Var1)
```

**Visualise** 

```{r viz, fig.width=12,fig.height=4}
p1 = DimPlot(allCells, reduction = "umap",label = F)
# p1 + scale_color_lancet() # doens't look good 
p2 = DimPlot(allCells, reduction = "umap",label = F)  + theme(legend.position = "none")
p3 = DimPlot(allCells, reduction = "umap",label = T) + theme(legend.position = "none")
CombinePlots(plots = list(p1,p2,p3),ncol = 3)
# save_plot(tikiR::figpath("allCells_clusters.pdf"),last_plot(),base_width=12,base_height=4)
# save_plot(tikiR::figpath("allCells_clusters_large.pdf"),last_plot(),base_width=15,base_height=5)
```

```{r}
# DimPlot(allCells, reduction = "umap",split.by = "group")  + theme(legend.position = "none") 
p1 = DimPlot(allCells, reduction = "umap",group.by = "group",pt.size = 0.05) + theme_void()+ theme(legend.position = "none") 
p2 = DimPlot(allCells, reduction = "umap",group.by = "pid",pt.size = 0.05) + theme_void()+ theme(legend.position = "none") 
CombinePlots(plots = list(p1,p2),ncol = 2,nrow=2) 
# save_plot(tikiR::figpath("Figure1_group-pid_noLabel.pdf"),last_plot(),base_width=3,base_height=3)
# allCells@meta.data %>% head
```


```{r viz-by-group}
DimPlot(allCells, reduction = "umap",label = F,split.by = "group")  + theme(legend.position = "none")
# save_plot(tikiR::figpath("Cluster-group-nolegend.pdf"),last_plot(),base_width=6,base_height=3)
DimPlot(allCells, reduction = "umap",label = F,split.by = "orig.ident",ncol=3) + theme(legend.position = "none")
# save_plot(tikiR::figpath("Cluster-pid-nolegend.pdf"),last_plot(),base_width=9,base_height=6)
```

Identify differentially expressed genes (DEG) of clusters, using MAST for statistical testing, with the function `FindAllMarkers`. This compares each one cluster against all other clusters. 

```{r identifyClusters, eval=FALSE, include=F}
# plan("multiprocess",workers=12) # parallelization
# # plan() 
# Idents(allCells)
# ?FindAllMarkers
markers <- FindAllMarkers(object = allCells, only.pos = T,test.use = "MAST")
beep()
saveRDS(markers,"data/output/20200629 allCells_FindAllMarkers.rds")
write.csv2(markers,"data/output/20200629 allCells_FindAllMarkers.csv",row.names = F)
rm(markers)
```

```{r clusterMarkers}
markers = readRDS("data/output/20200629 allCells_FindAllMarkers.rds")

# describe markers
markers %>% 
  group_by(cluster) %>% 
  filter(p_val_adj < 0.05) %>%
  top_n(3, avg_logFC) %>% 
  kable(caption = "Top 3 genes per cluster by log fold change")

# make heatmap based on top 10 genes per cluster by log fold change & significant
top10 <- markers %>% 
  group_by(cluster) %>% 
  filter(p_val_adj < 0.05) %>%
  top_n(10, avg_logFC)

# write.csv2(top10,"data/output/20200629 top10markers.csv",row.names = F)

DoHeatmap(object = allCells, features = top10$gene,
          label = TRUE,angle=0,size=3)  # adjustments for the plot

# save_plot(tikiR::figpath("allCells-heatmap_top10.pdf"),last_plot(),base_width=10, base_height=12)
```


<!-- ```{r include=F,eval=F} -->
<!-- avgExp <- AverageExpression(allCells, return.seurat = TRUE) -->
<!-- saveRDS(avgExp,"data/processed/allCells_avgExp.rds") -->
<!-- ``` -->



<!-- ```{r} -->
<!-- avgExp = readRDS("data/processed/allCells_avgExp.rds") -->
<!-- Idents(avgExp) = fct_relevel(Idents(avgExp),"0","1","4","2","11","3","9","5","8","12","7","14","6","10","13","15","16") -->
<!-- levels(Idents(avgExp)) -->

<!-- top10 %<>% -->
<!--   ungroup() %>% -->
<!--   mutate(cluster = cluster %>% as.factor %>% fct_relevel(levels(Idents(avgExp)))) %>% -->
<!--   arrange(cluster) -->

<!-- # allCells = ScaleData(allCells,assay="RNA") -->
<!-- DoHeatmap(object = avgExp, features = top10$gene,# assay = "RNA", -->
<!--           angle=0,size=3,draw.lines = F) # adjustments for the plot -->


<!-- # save_plot(tikiR::figpath("allCells_heatmap_top10_avgExp.pdf"),last_plot(),base_width=5, base_height=12) -->
<!-- rm(top10,avgExp) -->
<!-- ``` -->


```{r}
DefaultAssay(allCells) <- "RNA" 

mod = theme_cowplot() + theme(legend.position="none",axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank())

p1 = FeaturePlot(object = allCells, features = c("CD3G"))  
# save_plot(tikiR::figpath("CD3G_example.pdf"),last_plot(),base_width=4,base_height=3)

p1 = FeaturePlot(object = allCells, features = c("CD3G"))  + mod
p2 = FeaturePlot(object = allCells, features = c("CD4")) + mod
p3 = FeaturePlot(object = allCells, features = c("CD8A")) + mod
p4 = FeaturePlot(object = allCells, features = c("KLRD1")) + mod
p5 = FeaturePlot(object = allCells, features = c("CD79A")) + mod

p11 = FeaturePlot(object = allCells, features = c("IFNG"))  + mod
p12 = FeaturePlot(object = allCells, features = c("CCR7"))  + mod
p13 = FeaturePlot(object = allCells, features = c("SELL"))  + mod
p14 = FeaturePlot(object = allCells, features = c("FOXP3"))  + mod
p15 = FeaturePlot(object = allCells, features = c("FCGR3A"))  + mod


plot_grid(p1,p2,p3,p4,p5,p11,p12,p13,p14,p15,ncol=5)
# save_plot(tikiR::figpath("subsets.pdf"),last_plot(),base_width=16,base_height=7)
beep()
```


**Cluster identity**

```{r}
# gather extra data to distinguish similar clusters pairwise
markers.to.plot <- c("CD3D", "CREM", "CCR7", "SELL", "GIMAP5", "CACYBP", "GNLY", "NKG7", "CCL5", 
    "CD8A", "MS4A1", "CD79A", "MIR155HG", "NME1", "FCGR3A", "VMO1", "CCL2", "S100A9", "HLA-DQA1", 
    "GPR183", "PPBP", "GNG11", "HBA2", "HBB", "TSPAN13", "IL3RA", "IGJ")

DefaultAssay(allCells) <- "RNA"
DotPlot(allCells, features = rev(markers.to.plot), 
        cols = c("white", "#de2d26"),
        dot.scale = 5) + RotatedAxis()
DefaultAssay(allCells) <- "integrated"
rm(markers.to.plot)
```


```{r}
# write out the clusters so we can focus downstream analyses

clusters = data.frame(cluster = 0:16, 
           # inferred from top 3-10 genes per cluster
           cellType = c(
             # in sets of 5
             "CCR7+ T #1", "CCR7+ T #2", "ITGB+ IL7R+ T", "CD8", "IL7R+ T", 
             "NKG7+ CD8", "B naive", "Monocyte", "FCGR3A+ NK ", "ZNF683+ CD8",
             "B memory", "LINC02446+ CD8", "XCL1+ NK", "RBC/plt", "FCGR3A+ Monocyte", 
             # 14 could be non-classical monocyte
             "unclear", "pDC"
              ) 
           ) %>%
  mutate(cellType = cellType %>% as.character)
clusters %>% kable


# temporary df to match cell cluster and annotation
x = data.frame(cluster = allCells$seurat_clusters %>% as.character %>% as.numeric)
x %<>% left_join(clusters)
allCells$cellType = x$cellType

Idents(allCells) = "cellType"
DimPlot(allCells, reduction = "umap",label = T) + theme(legend.position = "none")
# save_plot(tikiR::figpath("allCells_clusters_Labelled.pdf"),last_plot(),base_width=6,base_height=6)

rm(x)
```



**Cluster proportions**

```{r}
allCells.list = data.frame(cluster = allCells$seurat_clusters, group = allCells$group,pid = allCells$orig.ident)
allCells.list$cellID = rownames(allCells.list)

allCells.list %<>%
  mutate(group = group %>% fct_relevel("transient","chronic"))

# calculate proportion of cells in each cluster per patient
proportions = allCells.list %>%
  add_count(pid,name = "N") %>%
  count(pid,cluster,N) %>%
  mutate(proportion=n/N)   

# add group data
proportions %<>% 
  left_join(allCells.list %>% count(pid,group) %>% select(pid,group),# fix to quickly access group data
             by="pid")

# proportions %<>%
  # left_join(clusters %>% mutate(cluster = cluster %>% as.character))

rm(allCells.list)
```

```{r}
p1 = proportions %>% 
  left_join(clusters %>% mutate(cluster = as.factor(as.character(cluster))),by="cluster")%>% 
  left_join(allClusters %>% rename(cluster = Cluster,TotalNCluster = Freq)) %>%
  mutate(cellType = as.factor(cellType) %>% fct_reorder(TotalNCluster)) %>%
  filter(cluster %in% 0:8) %>%
  ggplot(aes(cellType,proportion)) +
  geom_text(aes(y=0.25,label=TotalNCluster,hjust=0)) +
  geom_text(aes(y=0.3,label=cluster,hjust=0)) +
  stat_summary(geom="errorbar",aes(group=group,col=group),fun.data = mean_se,width=0.2) + #,position=position_dodge(width=0.8)) +
  stat_summary(geom="point",aes(group=group,col=group),fun.y = mean) +# ,position=position_dodge(width=0.8)) +#,position = "dodge") +
  coord_flip() +
  scale_y_continuous(position="right",breaks = c(0,0.1,0.2),labels=c(0,0.1,0.2),limits = c(0,0.3)) +
  scale_color_lancet() +
  theme(axis.ticks.y = element_blank(),axis.line.y=element_blank(),legend.position = "none",
        axis.line.x=element_blank()) +
  labs(x="",y="") 
  
p2 = proportions %>% 
  left_join(clusters %>% mutate(cluster = as.factor(as.character(cluster))),by="cluster")%>% 
  left_join(allClusters %>% rename(cluster = Cluster,TotalNCluster = Freq)) %>%
  mutate(cellType = as.factor(cellType) %>% fct_reorder(TotalNCluster)) %>%
  filter(cluster %in% 9:20) %>%
  ggplot(aes(cellType,proportion)) +
  geom_text(aes(y=0.06,label=TotalNCluster,hjust=0)) +
  geom_text(aes(y=0.09,label=cluster,hjust=0)) +
  stat_summary(geom="errorbar",aes(group=group,col=group),fun.data = mean_se,width=0.2) + #,position=position_dodge(width=0.8)) +
  stat_summary(geom="point",aes(group=group,col=group),fun.y = mean) +# ,position=position_dodge(width=0.8)) +#,position = "dodge") +
  coord_flip() +
  scale_y_continuous(position="right",breaks = c(0,0.025,0.05),labels=c(0,0.025,0.05),limits = c(0,0.1)) +
  scale_color_lancet() +
  theme(axis.ticks.y = element_blank(),axis.line.y=element_blank(),legend.position = "none",
        axis.line.x=element_blank()) +
  labs(x="",y="") 
  
p2

plot_grid(p1,p2,ncol=1,rel_heights = c(9/17,8/17),align = "v")
# save_plot(tikiR::figpath("Fig1_proportions_allCells.pdf"),last_plot() ,base_width=5,base_height=6)

```


```{r graph-proportions}

clusterLabels = clusters$cellType
names(clusterLabels) = clusters$cluster

# graph the proportions
proportions %>% 
  filter(cluster %in% 0:8) %>%
  ggplot(aes(group,proportion,col=group)) +
  stat_summary(geom="errorbar",width=0.2) +
  stat_summary(geom="point") +
  geom_point(alpha=0.5) +
  facet_wrap(~cluster,scales = "free_x",ncol=5,
              labeller = labeller(cluster = clusterLabels)) +
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        legend.position = "top",
        legend.title = element_blank()) +
  xlab("") +
  scale_color_lancet()

# save_plot(tikiR::figpath("proportions_allCells.pdf"),last_plot() + theme(legend.position = "none"),base_width=12,base_height=6)

proportions %>% 
  filter(cluster %in% 9:16) %>%
  ggplot(aes(group,proportion,col=group)) +
  stat_summary(geom="errorbar",width=0.2) +
  stat_summary(geom="point") +
  geom_point(alpha=0.5) +
  ylim(0,0.05) +
  facet_wrap(~cluster,scales = "free_x",ncol=5,
             labeller = labeller(cluster = clusterLabels)) +
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        legend.position = "top",
        legend.title = element_blank()) +
  xlab("") +
  scale_color_lancet()

# save_plot(tikiR::figpath("proportions_allCells_part2.pdf"),last_plot() + theme(legend.position = "none"),base_width=12,base_height=6)


```

```{r}
for (i in 0:16){
  print(paste0("Results for cluster ",i))
  test = proportions %>%
  filter(cluster==i) %$%
  t.test(proportion~group) 
  
  print(test)
}

rm(proportions,clusters,clusterLabels)
```

### Canonical marker expression heatmao


```{r}
allCells@active.ident

allCells.filtered = subset(allCells, idents = c("CCR7+ T #1", "CCR7+ T #2", "IL7R+ T", "ITGB+ IL7R+ T",   "CD8", "LINC02446+ CD8", "ZNF683+ CD8","NKG7+ CD8", "FCGR3A+ NK ", "XCL1+ NK", "B naive",  "B memory", "Monocyte", "FCGR3A+ Monocyte","pDC","RBC/plt", "unclear"))

?subset.Seurat
avgExp <- AverageExpression(allCells.filtered, return.seurat = TRUE)
avgExp$orig.ident


# selected T cell  features
# avgExp = ScaleData(avgExp)
genes = c("CCR7","SARAF", "IL7R","FOXP3","CD8A","ITGB1","IFNG","LAG3","HLA-DRB1","CXCR3",
       "IL2RA",
         "KLRD1","KLRC1", "GZMA","GZMB", "FCGR3A","XCL1","XCL2", 
           "CD19","CD79A","IGKC","JCHAIN","BANK","IGHA1", 
           "IL3RA","CLEC4A","LILRA4","IRF7","LILRB1", "CD74","LYZ"#, 
           #"TUBB1","PF4","SPARC" 
           )




# cluster re-assignment occurs, which re-assigns clustering in my_levels (assuming you have 12 clusters in total)
clusters %<>% mutate(cellType = as.factor(cellType) %>% fct_reorder(cluster)) 
my_levels <- levels(clusters$cellType)
#levels(clusters$cellType) %>% dput
my_levels = c("CCR7+ T #1", "CCR7+ T #2", "IL7R+ T", "ITGB+ IL7R+ T",   "CD8", "LINC02446+ CD8", "ZNF683+ CD8",
"NKG7+ CD8", "FCGR3A+ NK ", "XCL1+ NK", "B naive",  "B memory", "Monocyte", "FCGR3A+ Monocyte","pDC",
  "RBC/plt", "unclear" )

# Re-level object@ident
avgExp@active.ident <- factor(x = avgExp@active.ident, levels = my_levels)


DoHeatmap(avgExp, features = genes, size = 3, 
    draw.lines = F,raster = FALSE,group.bar = T)#,label = c(0:20) %>% as.character())

save_plot(tikiR::figpath("Figure1_heatmap.pdf"),last_plot(),base_width=7,base_height = 6)
```



# T cells 

**Clustering**

```{r}
Idents(allCells) = "seurat_clusters"
tCells = subset(allCells,idents = c(0,1,2,3,4,5,9,11))
tCells <- RunPCA(tCells, npcs = 50)
ElbowPlot(tCells,ndims = 50)
```

* Compute and store 10 PCs (elbow) by PCA using `RunPCA` using variable features for the Assay
* Dimensional reduction by UMAP using `RunUMAP` using PCA dimensional reduction as input, using the python package umap-learn
* Shared Nearest Neighbor (SNN) graph, determining k-nearest neighbors of each cell (k=20) using the 10 PC dimensions as input with `FindNeighbors`
* Identify clusters based on PCs, using the SNN, with the function `FindClusters` and a resolution of 0.8 (standard). The output is saved as `$seurat_clusters`

```{r cluster-tCells, eval=FALSE, include=F}
tCells <- RunPCA(tCells, npcs = 10) #recalc so we store only 10
tCells <- RunUMAP(tCells, reduction = "pca", dims = 1:10)
tCells <- FindNeighbors(tCells, reduction = "pca", dims = 1:10)
tCells <- FindClusters(tCells)

saveRDS(tCells,"data/processed/20200626 tCells_Clustered.rds")
rm(tCells)
```

Cells per cluster

```{r}
tCells = readRDS("data/processed/20200626 tCells_Clustered.rds")
table(tCells$seurat_clusters) %>% kable(caption="T cells per identified seurat cluster(all samples)")
```

**Visualise** 

```{r viz-tCell, fig.width=12,fig.height=4}
p1 = DimPlot(tCells, reduction = "umap",label = F)
# p1 + scale_color_lancet() # doens't look good 
p2 = DimPlot(tCells, reduction = "umap",label = F)  + theme(legend.position = "none")
p3 = DimPlot(tCells, reduction = "umap",label = T) + theme(legend.position = "none")
CombinePlots(plots = list(p1,p2,p3),ncol = 3)
rm(p1,p2,p3)
# save_plot(tikiR::figpath("tCells_clusters.pdf"),last_plot(),base_width=12,base_height=4)
# save_plot(tikiR::figpath("tCells_clusters_large.pdf"),last_plot(),base_width=15,base_height=5)
```

```{r viz-tCell-by-group}
DimPlot(tCells, reduction = "umap",label = F,split.by = "group")  + theme(legend.position = "none")
# save_plot(tikiR::figpath("tCells-Cluster-group-nolegend.pdf"),last_plot(),base_width=6,base_height=3)
DimPlot(tCells, reduction = "umap",label = F,split.by = "orig.ident",ncol=3) + theme(legend.position = "none")
# save_plot(tikiR::figpath("tCells-Cluster-pid-nolegend.pdf"),last_plot(),base_width=9,base_height=6)
```

Identify differentially expressed genes (DEG) of clusters, using MAST for statistical testing, with the function `FindAllMarkers`. This compares each one cluster against all other clusters. 

```{r identifyClusters-tCell, eval=FALSE, include=F}
plan("multiprocess",workers=12) # parallelization
plan() 
# Idents(allCells)
# ?FindAllMarkers
markers <- FindAllMarkers(object = tCells, only.pos = T,test.use = "MAST")
beep()
saveRDS(markers,"data/output/20200629 tCells_FindAllMarkers.rds")
write.csv2(markers,"data/output/20200629 tCells_FindAllMarkers.csv",row.names = F)
rm(markers)
```

```{r clusterMarkers-tCell}
markers = readRDS("data/output/20200629 tCells_FindAllMarkers.rds")

# describe markers
markers %>% 
  group_by(cluster) %>% 
  filter(p_val_adj < 0.05) %>%
  top_n(3, avg_logFC) %>% 
  kable(caption = "Top 3 significant genes per T cell cluster by log fold change")

# make heatmap based on top 10 genes per cluster by log fold change & significant
top10 <- markers %>% 
  group_by(cluster) %>% 
  filter(p_val_adj < 0.05) %>%
  top_n(10, avg_logFC)

# write.csv2(top10,"data/output/20200626 top10markers.csv",row.names = F)

#Idents(tCells) = "seurat_clusters"
# DoHeatmap(object = tCells, features = top10$gene, 
          # label = TRUE,angle=0,size=3)  # adjustments for the plot

avgExp <- AverageExpression(tCells, return.seurat = TRUE)
Idents(avgExp) = fct_relevel(Idents(avgExp),"5","4","0","1","3","8","10","2","7","9","6")
levels(Idents(avgExp))

top10 %<>%
  ungroup() %>%
  mutate(cluster = cluster %>% as.factor %>% fct_relevel(levels(Idents(avgExp)))) %>%
  arrange(cluster)

avgExp = ScaleData(avgExp,assay="RNA")
DoHeatmap(object = avgExp, features = top10$gene,# assay = "RNA",
          angle=0,size=3,draw.lines = F) # adjustments for the plot


save_plot(tikiR::figpath("tCells_heatmap_top10.pdf"),last_plot(),base_width=5, base_height=12)
rm(top10)
```

**Cluster identity**

```{r}
# gather extra data to distinguish similar clusters pairwise
markers.to.plot <- c("CD3D", "IL7R", "CCR7", "GZMA","CD8A","CD8B","CXCR3","GZMK","KLRD1","XCL1","KLRF1","EOMES","CX3CR1",
                     "CD4","FOXP3","CTLA4","IL2RA","CXCR6","KLRB1")

DefaultAssay(tCells) <- "RNA"
DotPlot(tCells, features = rev(markers.to.plot),
        cols = c("white", "#de2d26"),
        dot.scale = 5) + RotatedAxis()
DefaultAssay(tCells) <- "integrated"
rm(markers.to.plot)
```

```{r include=F,eval=F}
# targeted details by pairwise comparison

compareIdents = function(object, ident1,ident2,annotation){
    # Find markers that are conserved between the groups for cell identity 1 compared to cell identity 2
    conserved.markers = FindConservedMarkers(object,ident1,ident2,grouping.var = "group") 
    conserved.markers$gene = rownames(conserved.markers)
    
    # conserved.markers %<>%
      # filter(max_pval < 0.05)
    
    # Select 'avg_logFC' columns indices
    avg_logFC_columns <- grep(pattern = "avg_logFC", x = colnames(x = conserved.markers))
    
    # Compute mean 'avg_logFC'
    conserved.markers$mean_avg_logFC <- rowMeans(x = conserved.markers[avg_logFC_columns])
    
    ## For upregulated genes
    # Order 'conserved.markers' by 'mean_avg_logFC'
    conserved.markers <- conserved.markers[order(conserved.markers$mean_avg_logFC, decreasing = TRUE), ]
    
    # Get top 10 genes
    up = rownames(x = conserved.markers[1:10,])
    
    ## For downregulated genes
    # Order 'conserved.markers' by 'mean_avg_logFC'
    conserved.markers <- conserved.markers[order(conserved.markers$mean_avg_logFC, decreasing = FALSE), ]
    
    # Get top 10 genes
    down = rownames(x = conserved.markers[1:10,])
    
    out = conserved.markers %>% 
      filter(gene %in% up | gene %in% down) %>% 
      mutate(ident.1 = ident1, 
             ident.2 = ident2) %>%
      select(gene, ident.1,ident.2, mean_avg_logFC,max_pval) %>%
      arrange(desc(mean_avg_logFC))
    
    # log the output
    write.csv2(out,paste0("data/output/conservedMarkers_",annotation,"_ident",ident1,"_vs",ident2,".csv"))
    print(paste0("wrote results to data/output/conservedMarkers_",annotation,"_ident",ident1,"_vs",ident2,".csv"))
    
    return(out)
}

compareIdents(tCells,2,10,annotation="tCell")
compareIdents(tCells,2,7,annotation="tCell")

compareIdents(tCells,7,3,annotation="tCell")

compareIdents(tCells,1,0,annotation="tCell")
compareIdents(tCells,5,0,annotation="tCell")
compareIdents(tCells,5,1,annotation="tCell")
compareIdents(tCells,0,4,annotation="tCell")

compareIdents(tCells,3,1,annotation="tCell")
compareIdents(tCells,8,3,annotation="tCell")

compareIdents(tCells,6,9,annotation="tCell")
compareIdents(tCells,7,9,annotation="tCell")
```


<!-- ```{r} -->
<!-- # write out the clusters so we can focus downstream analyses -->

<!-- clusters = data.frame(cluster = 0:10,  -->
<!--            # inferred from top 3-10 genes per cluster -->
<!--            cellType = c( -->
<!--              # in sets of 5 -->
<!--              "0","1","CD8 LINC02446+ #1","3","T-B doublet?", -->
<!--              "SOX4","GZMB+", "ZNF683+","8","GZMK+", -->
<!--              "CD8 LINC02446+ #2" -->
<!--               )  -->
<!--            ) %>% -->
<!--   mutate(cellType = cellType %>% as.character) -->
<!-- clusters %>% kable -->


<!-- # temporary df to match cell cluster and annotation -->
<!-- x = data.frame(cluster = tCells$seurat_clusters %>% as.character %>% as.numeric) -->
<!-- x %<>% left_join(clusters) -->
<!-- tCells$cellType = x$cellType -->

<!-- Idents(tCells) = "cellType" -->
<!-- DimPlot(tCells, reduction = "umap",label = T) + theme(legend.position = "none") -->
<!-- # save_plot(tikiR::figpath("allCells_clusters_Labelled.pdf"),last_plot(),base_width=6,base_height=6) -->

<!-- rm(x) -->
<!-- ``` -->

```{r}
# avgExp = AverageExpression(tCells)
# head(avgExp[["RNA"]])

# Idents(tCells)
# Idents(tCells) = paste("cluster",Idents(tCells),sep="")

avgExp <- AverageExpression(tCells, return.seurat = TRUE)
avgExp$orig.ident


# selected T cell  features
# DefaultAssay(avgExp) = "integrated"
# DefaultAssay(avgExp) = "RNA"
# avgExp = ScaleData(avgExp)
genes =  c("CCR7","SELL","LDHB","LRRN3",  #naive, our data (PCA)
           "FHIT","TCF7", "FCGRT",
           "JUN","JUNB", "FOS", "CD69", # early activation
           "CD8A","CTSW",
           "NKG7","CST7", "CCL5","KLRD1","GZMH","PRF1", "FCGR3A", #CD8
           "ZNF683","ITGB1", "S100A4","ANXA1","KLF6","LGALS3","CYTOR", #activ
           "FOXP3","IL2RA","TIGIT","CTLA4","IL32", "PLCL1","SHMT2","C21orf91" #Treg by Cuadrado Immunity 2018
        )

DoHeatmap(avgExp, features = genes, size = 3, 
    draw.lines = F,raster = FALSE,group.bar = T)
# save_plot(tikiR::figpath("tCells_heatMap_Cannonical.pdf"),last_plot(),base_width =5,base_height=6)
```

```{r}
genes = c("CD8A","CCL5", "CCR7","ITGB1",
          "IL2RA","CTLA4")

FeaturePlot(tCells,features=genes,min.cutoff = "q20")
```


**Cluster proportions**

```{r}
tCells.list = data.frame(cluster = tCells$seurat_clusters, group = tCells$group,pid = tCells$orig.ident)
tCells.list$cellID = rownames(tCells.list)

tCells.list %<>%
  mutate(group = group %>% fct_relevel("transient","chronic"))

# calculate proportion of cells in each cluster per patient
proportions = tCells.list %>%
  add_count(pid,name = "N") %>%
  count(pid,cluster,N) %>%
  mutate(proportion=n/N)   

# add group data
proportions %<>% 
  left_join(tCells.list %>% count(pid,group) %>% select(pid,group),# fix to quickly access group data
             by="pid")

# proportions %<>%
  # left_join(clusters %>% mutate(cluster = cluster %>% as.character))


```


```{r graph-proportions-tCells}

clusterLabels = 0:10 #clusters$cellType
names(clusterLabels) = 0:10 #clusters$cluster

# graph the proportions
proportions %>% 
  filter(cluster %in% 0:10) %>%
  ggplot(aes(group,proportion,col=group)) +
  stat_summary(geom="errorbar",width=0.2) +
  stat_summary(geom="point") +
  geom_point(alpha=0.5) +
  facet_wrap(~cluster,scales = "free_x",ncol=5,
              labeller = labeller(cluster = clusterLabels)) +
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        legend.position = "top",
        legend.title = element_blank()) +
  xlab("") +
  scale_color_lancet()

# save_plot(tikiR::figpath("proportions_tCells.pdf"),last_plot() + theme(legend.position = "none"),base_width=12,base_height=9)

```

```{r}
for (i in 0:10){
  print(paste0("Results for cluster ",i))
  test = proportions %>%
  filter(cluster==i) %$%
  t.test(proportion~group)
  
  print(test)
}

rm(clusters,clusterLabels)
```

**Cum proportions**

CD8 vs CD4

```{r}
subtypes = proportions %>%
  select(-proportion) %>%
  mutate(type = case_when(
    cluster %in% c("2","10", "7","9","6") ~ "CD8",
    TRUE ~ "CD4"
  )) %>%
  group_by(pid,N,type) %>%
  summarise(sum_type = sum(n)) %>%
  # rename(sum_type = n) %>%
  mutate(proportion=sum_type/N)  

# add group data
subtypes %<>% 
  left_join(tCells.list %>% count(pid,group) %>% select(pid,group),# fix to quickly access group data
             by="pid")

subtypes %>%
  ggplot(aes(group,proportion,col=group)) +
  stat_summary(geom="errorbar",width=0.2) +
  stat_summary(geom="point") +
  geom_point(alpha=0.5) +
  facet_wrap(~type,scales = "free_x",ncol=5)+
              # labeller = labeller(cluster = clusterLabels)) +
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        legend.position = "top",
        legend.title = element_blank()) +
  xlab("") +
  scale_color_lancet()

# save_plot(tikiR::figpath("proportions_CD8.pdf"),last_plot() + theme(legend.position = "none"),base_width=6,base_height=3)
```

CCR7+ vs negative  CD4 

```{r}
subtypes = proportions %>%
  select(-proportion) %>%
  filter(!(cluster %in% c("2","10", "7","9","6")))%>%
  mutate(CCR7 = case_when(
    cluster %in% c("3","8") ~ "CCR7-",
    TRUE ~ "CCR7+"
  )) %>%
  group_by(pid,N,CCR7) %>%
  summarise(sum_type = sum(n)) %>%
  group_by(pid)%>%
  mutate(N=sum(sum_type)) %>%
  # rename(sum_type = n) %>%
  mutate(proportion=sum_type/N)  

subtypes %>%
  group_by(pid)%>%
  mutate(N=sum(sum_type))

# add group data
subtypes %<>% 
  left_join(tCells.list %>% count(pid,group) %>% select(pid,group),# fix to quickly access group data
             by="pid")

subtypes %>%
  ggplot(aes(group,proportion,col=group)) +
  stat_summary(geom="errorbar",width=0.2) +
  stat_summary(geom="point") +
  geom_point(alpha=0.5) +
  facet_wrap(~CCR7,scales = "free_x",ncol=5)+
              # labeller = labeller(cluster = clusterLabels)) +
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        legend.position = "top",
        legend.title = element_blank()) +
  xlab("") +
  scale_color_lancet()

# save_plot(tikiR::figpath("proportions_CCR7_CD4.pdf"),last_plot() + theme(legend.position = "none"),base_width=6,base_height=3)
```

CCR7+ vs negative CD8 

```{r}
subtypes = proportions %>%
  select(-proportion) %>%
  filter((cluster %in% c("2","10", "7","9","6")))%>%
  mutate(CCR7 = case_when(
    cluster %in% c("7","9","6") ~ "CCR7-",
    TRUE ~ "CCR7+"
  )) %>%
  group_by(pid,N,CCR7) %>%
  summarise(sum_type = sum(n)) %>%
  group_by(pid)%>%
  mutate(N=sum(sum_type)) %>%
  mutate(proportion=sum_type/N)  

# add group data
subtypes %<>% 
  left_join(tCells.list %>% count(pid,group) %>% select(pid,group),# fix to quickly access group data
             by="pid")

subtypes %>%
  ggplot(aes(group,proportion,col=group)) +
  stat_summary(geom="errorbar",width=0.2) +
  stat_summary(geom="point") +
  geom_point(alpha=0.5) +
  facet_wrap(~CCR7,scales = "free_x",ncol=5)+
              # labeller = labeller(cluster = clusterLabels)) +
  theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        legend.position = "top",
        legend.title = element_blank()) +
  xlab("") +
  scale_color_lancet() +
  ylim(0.2,0.8)

# save_plot(tikiR::figpath("proportions_CCR7_CD8.pdf"),last_plot() + theme(legend.position = "none"),base_width=6,base_height=3)
```

```{r}
rm(tCells.list,proportions,subtypes)
```


<!-- Better define Treg -->


<!-- ```{r} -->
<!-- rm(tCells) -->
<!-- allCells = readRDS("data/processed/20200626 allCells_Clustered.rds") -->
<!-- Idents(allCells) = "seurat_clusters" -->
<!-- tCells = subset(allCells,idents = c(0,1,2,3,4,5,9,11)) -->

<!-- tCells = RunPCA(tCells, features=c("FOXP3","CTLA4","IL32","TIGIT","IL2RA","FOXP3", #Treg -->
<!--                            "CD3D", "CD8A","CD8B","CTSW",# CD8 -->
<!--                            "NKG7","CST7","CCL5","KLRD1","GMZH","ZNF683","ITGB1","S100A4", #activ -->
<!--                            "IL7R","CD69","CCR7","SELL","LDHB","FHIT","LRRN3"#,#Tn -->
<!--                           # "CXCR3","CX3CR1","EOMES" -->
<!--                            )) -->

<!-- tCells <- RunUMAP(tCells, reduction = "pca", dims = 1:10) -->
<!-- tCells <- FindNeighbors(tCells, reduction = "pca", dims = 1:10) -->
<!-- tCells <- FindClusters(tCells,resolution = 0.3) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- DimPlot(tCells) -->
<!-- FeaturePlot(tCells,c("FOXP3","CTLA4","CD8A","CCR7"),min.cutoff = "q20") -->
<!-- ``` -->


# comparison to flow cytometry analyses

```{r}
CD8_clusters = c("2","10", "7","9","6") 
'%!in%' <- function(x,y)!('%in%'(x,y))

Idents(tCells)
tCells@meta.data

CD4 = subset(tCells, (seurat_clusters %!in% CD8_clusters))
CD8 = subset(tCells, (seurat_clusters %in% CD8_clusters))
```

```{r}
Assays(tCells)
DefaultAssay(tCells) = "RNA"
FeatureScatter(CD4, "CD27","CCR7")
FeatureScatter(CD4, "SELL","CCR7")
```


```{r}
allCells@active.ident

avgExp <- AverageExpression(CD4, return.seurat = TRUE)

# selected T cell  features
# avgExp = ScaleData(avgExp)
genes = c("CCR7", "CD27", "SELL", "LEF1", "TCF7","GNLY","GZMB","GZMA","PRF1","SLAMF7","CX3CR1","ITGB1" )

# cluster re-assignment occurs, which re-assigns clustering in my_levels (assuming you have 12 clusters in total)
# clusters %<>% mutate(cellType = as.factor(cellType) %>% fct_reorder(cluster)) 
my_levels <- levels(avgExp@active.ident)
my_levels %>% dput
my_levels = c("4", "5", "0", "1", "3","8")

# Re-level object@ident
avgExp@active.ident <- factor(x = avgExp@active.ident, levels = my_levels)


DoHeatmap(avgExp, features = genes, size = 3, 
    draw.lines = F,raster =     FALSE,group.bar = T)#,label = c(0:20) %>% as.character())

save_plot(tikiR::figpath("tCell_heatmap-TnTcmTeff.pdf"),last_plot(),base_width=4.5,base_height = 4)
```


```{r outro}
devtools::session_info()
```

